The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-11 Independence_CH/related.tex
On ZFC_in_HOL

This faithfulness to set-theoretical practice does not come for
free. Recursive constructions and inductive definitions are far easier
to perform in the Calculus of Inductive Constructions on which Lean
(and Coq) are based, and in Isabelle/ZF are rather cumbersome. Also, a
typed discipline provides aid to write succintly and many assumptions
are satisfied by mere notation. To be clear, those benefits come from
doing set theory in a non set-theoretical meta-language. On the other
hand, Isar proofs, as the one shown in
Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
than the language of tactics of Lean.
%% Comment on ZFC_in_HOL and other developments using it (mostly by
%% Paulson: Ramsey theory, (ordinal) partition theorems)

>>>>

This faithfulness to set-theoretical practice does not come for
free. Recursive constructions and inductive definitions are far easier
to perform in the Calculus of Inductive Constructions on which Lean
(and Coq) are based, and in Isabelle/ZF are rather cumbersome. Also, a
typed discipline provides aid to write succintly and many assumptions
are satisfied by mere notation. To be clear, those benefits come from
doing set theory in a non set-theoretical meta-language. On the other
hand, Isar proofs, as the one shown in
Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
than the language of tactics of Lean.

A sweet spot combining the best of both worlds is to be found on
developments in Isabelle/HOL based on the AFP entry
\session{ZFC\_in\_HOL} by Paulson \cite{ZFC_in_HOL-AFP}. There is a
range of results in combinatorics and other set-theoretical material
that was swiftly formalized in this setting: Nash-Williams theorem and
Larson's $\forall k\ \omega^{\omega}\longrightarrow(\omega^\omega,k)$
\cite{doi:10.1080/10586458.2021.1980464}, Design Theory
\cite{10.1007/978-3-030-81097-9_1}, and Wetzel's problem
\cite{2022arXiv220503159P}; this last paper describes in a brief and
clear way the conveniences of the interaction with Isabelle/HOL
(which, in this case, is also paid in consistency-strength currency
\cite[Sect.~3]{DBLP:conf/ictac/Obua06}).
